The imprint of global magma oceans on exoplanet demographics

### 2023-04-09 src/tex/ms.tex
6.2.2.

An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from a similar mass reservoir~\citep{2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets \citep{2020ChEG...80l5594F}, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.

@@

\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies sufficient amounts of water vapor available in the atmosphere to act as greenhouse gas, but already about ~10--20 bar of water vapour -- a minor fraction of one Earth ocean (the lower limit of water on Earth) -- is enough to sustain sufficiently high surface temperatures to keep the surface molten, and thus in a magma ocean stage \citep{Boukrouche2021,2021JGRE..12606711L}. The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition. From a formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome of formation \citep{2019PNAS..116.9723Z,2020A&A...643L...1V}. But while ``typically'' considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System \citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation. Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amount of water inside of the planet during the magma ocean phase \citep{2018SSRv..214...76I,2021ApJ...909L..22K,2022NatAs...6.1296K}. Enhanced equilibration between the core, mantle, and atmospehre during magma ocean evolution of rocky exoplanets further enhances this process \citep{2021ApJ...914L...4L,Schlichting2022}. Therefore, even nominally dry planets generate substantial amounts of water during formation and early evolution. Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M dwarf systems \citep{2022ApJ...938L...3L}.

@@

\subsubsection{False positive scenarios}\label{sec:dis_falsepositive}
Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.

@@

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{Venturini2020,Burn2021,Schlecker2021,Schlecker2021b}, systems of all ages can be affected by this dilution.

>>>>

An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets \citep{2020ChEG...80l5594F}, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.

@@

\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies sufficient amounts of water vapor available in the atmosphere to act as greenhouse gas, but already about ~10--20 bar of water vapour -- a minor fraction of one Earth ocean (the lower limit of water on Earth) -- is enough to sustain sufficiently high surface temperatures to keep the surface molten, and thus in a magma ocean stage \citep{Boukrouche2021,2021JGRE..12606711L}. The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition. From a formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome of formation \citep{2019PNAS..116.9723Z,2020A&A...643L...1V}. But while ``typically'' considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System \citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation. Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amount of water inside of the planet during the magma ocean phase \citep{2018SSRv..214...76I,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}. Enhanced equilibration between the core, mantle, and atmospehre during magma ocean evolution of rocky exoplanets further enhances this process \citep{2021ApJ...914L...4L,Schlichting2022}. Therefore, even nominally dry planets generate substantial amounts of water during formation and early evolution. Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M dwarf systems \citep{2022ApJ...938L...3L}.

@@

\subsubsection{False positive scenarios}\label{sec:dis_falsepositive}
Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{2012ApJ...753...66I,Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates. For example, the K2-3 system shows tentative evidence for a break in the planet radii, contrary to what would be expected from atmospheric escape \citep{2022AJ....164..172D}.

@@

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{2019NatAs...3..307L,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.z
